Cross-validation of regression models: mixed model examples (edited v3)

### 2024-08-14 scratch/JSS-submission/Fox-Monette-cv-package.Rmd
small copy edits and one minor code change in code to reduce white space under Figure 5.

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0003-0076-5532
    affiliation: 'York University'
    address: |
      | Toronto, Ontario, Canada
    email: \email
    url: http://blackwell.math.yorku.ca/gmonette
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validating Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validating Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validating Regression Models"
abstract: >
  We describe the \pkg{cv} package, which implements cross-validation for standard \proglang{R} regression models through a uniform and simple to use generic function. Methods for the \code{cv()} function are provided for many commonly employed classes of statistical models, including mixed-effects models, and it is straightforward to extend the \pkg{cv} package by writing methods for additional classes of models. The \code{cv()} function also can cross-validate complex model-selection procedures, such as those that include variable transformations, predictor selection, and selection among competing models. The \code{cv()} function generally supports parallel computations, and the supplied methods for linear and generalized linear models take advantage of computational efficiencies.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv.bib"]
---

@@

Cross-validation speaks to both of these issues. In CV, the data are randomly divided as equally as possible into several, say $k$, parts, called "folds." The statistical model is fit $k$ times, leaving each fold out in turn. Each fitted model is then used to predict the response variable for the cases in the omitted fold. A CV criterion (also termed a "cost" or "loss" measure), such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values. In the extreme, $k = n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."

Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error. The $n$ regression models are highly statistically dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has larger variance than if the predictions from the models fit to the $n$ data sets were independent.

@@

The final section of the paper (Section \ref{computational-notes}) describes some computational details, including efficient CV computations for linear and generalized-linear models. and computation of bias-corrected CV criteria.

@@

We'll focus on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure \ref{fig:mpg-horsepower-polynomials} (a). The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-polynomials} (a). The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the degree-5 polynomial may over-fit the data by chasing one or two relatively high \code{mpg} values at the right (but see the CV results reported below).

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here. Consider, for example the scatterplot for log-transformed \code{mpg} and \code{horsepower}, produced by \code{plot(mpg ~ horsepower, data=Auto, log="xy")} (execution of which is left to the reader). We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.

@@

We illustrate with the polynomial regression models of varying degree for the \code{Auto} data, beginning by fitting and saving the 10 models:
```{r polyomial-models}
mlist <- vector(10, mode = "list")
for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
names(mlist) <- paste0("m.", 1:10)
mlist[2]
```
where, for example \code{mlist[2]} is the quadratic fit.

@@

In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-polynomials} (b) on page \pageref{fig:mpg-horsepower-polynomials}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.

@@

```{r cross-validation-data,include=FALSE,cache=TRUE}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })
dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r cross-validation-data-plot,echo=FALSE}
#| out.width = "75%",
#| fig.height = 3,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(clusters + cases ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', columns = 2, cex = 0.7), 
       type = 'b', ylab = list('CV criterion (MSE)', cex = .7),
       xlab = '',
       par.strip.text = list(cex = 0.7),
       layout= c(Nv, 1),
       par.settings =
         list(superpose.line=list(lty = c(1, 2), lwd = 1),
              superpose.symbol=list(cex = 0.7)),
       scales = list(y = list(log = TRUE, cex = 0.7), 
                     x = list(alternating = F, rot = 45, cex = 0.7))) 
```
In summary, when between-cluster variation is relatively large, the model \code{~ 1 + x}, with \code{x} alone and without the contextual mean \code{xm} of \code{x}, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. In all of our examples, the model \code{~ 1 + x + xm}, which includes both \code{x} and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models (in Figure \ref{fig:plot-fits-combined} on page \pageref{fig:plot-fits-combined}), and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

@@

And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore-cr-plots}, on page \pageref{fig:Auto-explore-cr-plots}, and Figure \ref{fig:Auto-transformed-cr-plots}:

@@

To cross-validate this multinomial-logit model we need an appropriate cost criterion. None of the criteria in the \pkg{cv} package will do---for example, \code{mse()} is appropriate only for a numeric response. The \code{BayesRule()} criterion, also supplied by \pkg{cv}, which is for a binary response, comes close:
```{r BayesRule}
BayesRule
```
After doing some error checking, \code{BayesRule()} rounds the predicted proabability \code{yhat} of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction, and then reports the proportion of incorrect predictions. Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, enabling the computation of bias correction and confidence intervals (as discussed in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).

>>>>

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0003-0076-5532
    affiliation: 'York University'
    address: |
      | Toronto, Ontario, Canada
    email: \email
    url: http://blackwell.math.yorku.ca/gmonette
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validating Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validating Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validating Regression Models"
abstract: >
  We describe the \pkg{cv} package, which implements cross-validation for standard \proglang{R} regression models through a uniform and simple to use generic function. Methods for the \code{cv()} function are provided for many commonly employed classes of statistical models, including mixed-effects models, and it is straightforward to extend the \pkg{cv} package by writing methods for additional classes of models. The \code{cv()} function can also cross-validate complex model-selection procedures, such as those that include variable transformations, predictor selection, and selection among competing models. The \code{cv()} function generally supports parallel computations, and the supplied methods for linear and generalized linear models take advantage of computational efficiencies.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
    extra_dependencies: "subfig"
bibliography: ["cv.bib"]
---

@@

Cross-validation speaks to both of these issues. In CV, the data are randomly divided as equally as possible into several, say $k$, parts, called "folds." The statistical model is fit $k$ times, leaving each fold out in turn. Each fitted model is then used to predict the response variable for the cases in the omitted fold. A CV criterion (also termed a "cost" or "loss" measure), such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values. In the extreme, $k$ equals $n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."

Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of the prediction error with a model fit to $n$ cases. The $n$ regression models are highly statistically dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has larger variance than if the predictions from the models fit to the $n$ data sets were independent.

@@

The final section of the paper (Section \ref{computational-notes}) describes some computational details, including efficient CV computations for linear and generalized-linear models, and computation of bias-corrected CV criteria.

@@

We'll focus on the relationship of \code{mpg} (miles per gallon) to \code{horsepower}, as displayed in Figure\ \ref{fig:mpg-horsepower-polynomials} (a). The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in Figure \ref{fig:mpg-horsepower-polynomials} (a). The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the degree-5 polynomial may over-fit the data by chasing one or two relatively high \code{mpg} values at the right (but see the CV results reported below).

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is not the best choice here. Consider, for example, the scatterplot for log-transformed \code{mpg} and \code{horsepower}, produced by \code{plot(mpg $\sim$ horsepower, data=Auto, log="xy")} (execution of which is left to the reader). We revisit the \code{Auto} data in Section \ref{cross-validating-model-specification}.

@@

We illustrate with the polynomial regression models of varying degree for the \code{Auto} data, beginning by fitting and saving the 10 models:
```{r polyomial-models}
mlist <- vector(10, mode = "list")
for (p in 1:10) mlist[[p]] <- lm(mpg ~ poly(horsepower, p), data = Auto)
names(mlist) <- paste0("m.", 1:10)
mlist[2]
```
where, for example, \code{mlist[2]} is the quadratic fit.

@@

In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model (cf., Figure \ref{fig:mpg-horsepower-polynomials} (b) on page \pageref{fig:mpg-horsepower-polynomials}), except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting.
\newpage

@@

```{r cross-validation-data,include=FALSE,cache=TRUE}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })
dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r cross-validation-data-plot,echo=FALSE}
#| out.width = "75%",
#| fig.height = 3,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(clusters + cases ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', columns = 2, cex = 0.7), 
       type = 'b', ylab = list('CV criterion (MSE)', cex = .7),
       xlab = '',
       par.strip.text = list(cex = 0.7),
       layout= c(Nv, 1),
       par.settings =
         list(layout.heights = list(bottom.padding = 0),
              superpose.line = list(lty = c(1, 2), lwd = 1),
              superpose.symbol = list(cex = 0.7)),
       scales = list(y = list(log = TRUE, cex = 0.7), 
                     x = list(alternating = F, rot = 45, cex = 0.7))) 
```
In summary, when between-cluster variation is relatively large, the model \code{~ 1 + x}, with \code{x} alone and without the contextual mean \code{xm} of \code{x}, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. In all of our examples, the model \code{~ 1 + x + xm}, which includes both \code{x} and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models (in Figure \ref{fig:plot-fits-combined} on page \pageref{fig:plot-fits-combined}), and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

@@

And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-explore-cr-plots}, on page \pageref{fig:Auto-explore-cr-plots}, and Figure \ref{fig:Auto-transformed-cr-plots}):

@@

To cross-validate this multinomial-logit model we need an appropriate cost criterion. None of the criteria in the \pkg{cv} package will do---for example, \code{mse()} is appropriate only for a numeric response. The \code{BayesRule()} criterion, also supplied by \pkg{cv}, which is for a binary response, comes close:
```{r BayesRule}
BayesRule
```
After doing some error checking, \code{BayesRule()} rounds the predicted probability \code{yhat} of a 1 ("success") response in a binary regression model to 0 or 1 to obtain a categorical prediction, and then reports the proportion of incorrect predictions. Because the Bayes's rule criterion is an average of casewise components (as, e.g., is the MSE), a \code{"casewise loss"} attribute is attached to the result, enabling the computation of bias correction and confidence intervals (as discussed in Section \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}).
